Analyzing the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions

### 2024-11-29 paper/paper.qmd
finalize paper and complete datasheet

---
title: "Analyzing the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
subtitle: "My subtitle if needed"
author: 
  - Kevin Cai
thanks: "Code and data are available at: [https://github.com/kevicai/toronto-healthcare-outbreak-prediction](https://github.com/kevicai/toronto-healthcare-outbreak-prediction)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
execute: 
  cache: true
bibliography: references.bib
---

@@

Outbreaks of infectious diseases in healthcare institutions are a significant challenge, particularly in hospitals, long-term care homes, and retirement homes, where vulnerable populations are most at risk. Understanding the factors that influence the duration of these outbreaks is important for improving preparedness, response times, and resource allocation. Previous studies have looked at factors affecting outbreak durations, but more research is needed, especially in Toronto’s healthcare system. This paper analyzes the key factors that influence the duration of outbreaks in Toronto healthcare institutions, focusing on outbreak setting, causative agents, and the timing of outbreaks.

This study uses a Bayesian negative binomial regression model to predict outbreak durations in Toronto's healthcare facilities. The model includes factors like the type of healthcare institution (outbreak setting), the causative agent, and the month the outbreak occurred. This research aims to identify which factors contribute most to variation in outbreak durations. It provides actionable information for public health officials and healthcare managers to improve their response strategies.

The results of this analysis help manage outbreaks, optimize patient care, and reduce the strain on healthcare facilities. Identifying high-risk settings or times of year when outbreaks last longer can improve resource allocation and response planning. This study also adds to the literature on healthcare outbreaks, with a focus on Toronto’s healthcare system.

@@

This report uses the Outbreaks in Toronto Healthcare Institutions dataset, contains data from January 2016 to November 2024. The dataset is provided by Toronto Public Health, through City of Toronto Open Data Portal [@datasource]. The dataset tracks reported outbreaks of gastroenteric and respiratory illnesses in Toronto healthcare institutions and contains detailed information on outbreak settings, causative agents, and outbreak durations. Following the principles from Telling Stories with Data [@textbook], we examine how the characteristics of outbreaks, such as the type of healthcare institution, the causative agent, and the month the outbreak began, influence their durations. A sample of the cleaned dataset is shown in @tbl-cleaned.

@@

### Outbreak Setting
The Outbreak Setting variable is categorical and identifies the type of healthcare institution where the outbreak occurred, such as hospitals, long-term care homes (LTCH), or retirement homes. It provides insights into the environments most affected by outbreaks.

@@

### Causative Agent
The Causative Agent variable is categorical and reflects the infectious agents responsible for outbreaks. While the original dataset contains 55 agents, they are grouped into seven broader categories to simplify the analysis and enhance interpretability.

@@

### Month Outbreak Began
The Month variable is categorical and records the calendar month when each outbreak started as a name (e.g., Jan, Feb). It reflects seasonal trends and potential patterns in infection rates. This variable is extracted from the date where each outbreak began from the original dataset and converted from a number to the corresponding month name.

@@

# Model

## Model Overview

@@

## Model Setup

@@

## Model Selection

@@

## Model Diagnostics and Validation

@@

# Results

To determine the reference levels for the categorical data outbreak settings, causative agent, and month, the model automatically selects the first category of each factor based on alphabetical orders of the category names. The intercept represents that the baseline outbreak duration is 2.257 days when all predictors are at their reference levels. @fig-90-interval shows that the

## Outbreak Setting
Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:
- Hospital-Chronic Care (0.047): Outbreaks in this setting last 0.047 days longer than Hospital-Acute Care.
- Hospital-Psychiatric (-0.080): Outbreaks in psychiatric hospitals last 0.080 days shorter than Hospital-Acute Care.
- LTCH (Long-Term Care Homes) (0.465): Outbreaks in LTCHs last 0.465 days longer than in Hospital-Acute Care.
- Retirement Home (0.343): Outbreaks in retirement homes last 0.343 days longer than in Hospital-Acute Care.

## Causative Agent
Outbreak setting is referenced based on "Hospital-Acute Care", we compare other categories with it to evaluate how each category influence the model:

Num. Obs. (3416): Based on 3416 observations.
Algorithm (sampling): The model used Bayesian sampling for estimation.

@@

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\appendix

@@

# Additional data details

# Model details {#sec-model-details}

## Outcome Variable Variance and Mean

## Posterior predictive check

@@

## Leave-One-Out (LOO) Cross Validation (CV) Comparison

@@

## Model summary

@@

## Diagnostics

>>>>

---
title: "Analyzing the Factors Influencing Outbreak Duration in Toronto Healthcare Institutions"
author: 
  - Kevin Cai
thanks: "Code and data are available at: [https://github.com/kevicai/toronto-healthcare-outbreak-prediction](https://github.com/kevicai/toronto-healthcare-outbreak-prediction)."
date: today
date-format: long
abstract: "This study investigates the factors influencing the duration of infectious disease outbreaks in healthcare institutions in Toronto, focusing on the role of outbreak setting, causative agent, and seasonal variations. Using a Bayesian negative binomial regression model, the analysis identifies significant drivers of outbreak duration and highlights the importance of targeted preparedness in settings like Long-Term Care Homes (LTCHs) and Retirement Homes. The findings suggest that COVID-19 has a substantial impact on outbreak duration compared to other respiratory viruses. This research underscores the need for robust healthcare infrastructure and adaptive strategies to manage future healthcare outbreaks effectively."
format: pdf
number-sections: true
execute: 
  cache: true
bibliography: references.bib
---

@@

Infectious disease outbreaks in healthcare settings present significant challenges, particularly in hospitals, long-term care homes, and retirement homes, where vulnerable populations are at highest risk [@torontoActiveOutbreaks]. These environments house individuals who are more susceptible to infections, including the elderly, individuals with compromised immune systems, and those with underlying health conditions. In Toronto, public health authorities are required to respond to and manage outbreaks of gastroenteric and respiratory illnesses in these institutions [@torontoActiveOutbreaks]. Efficient management of such outbreaks is essential to prevent widespread illness and ensure the safety of patients, staff, and visitors. Understanding the factors that influence the duration of these outbreaks is important for enhancing preparedness, minimizing response times, and optimizing resource allocation.

The factors influencing outbreak duration can vary significantly depending on the healthcare setting, the causative agent of the illness, and the timing of the outbreak. In nursing homes, for example, the close living conditions, shared spaces, and frequent interactions between residents, staff, and visitors contribute to the spread of infections. In hospitals, the high turnover of patients and healthcare workers, coupled with the need for intensive medical care, can also worsen the problem. Additionally, the characteristics of specific pathogens can influence transmission dynamics and outbreak duration. For instance, the concentration of SARS-CoV-2 in patient secretions peaks approximately 10 days after symptom onset, influencing transmission patterns and the length of outbreaks in healthcare facilities [@sarsOnHealthcare]. Understanding why some outbreaks persist longer than others requires identifying the drivers of these durations. While previous studies have examined factors affecting outbreak durations, more research is needed, especially within the context of Toronto's healthcare system. Toronto, with its large and diverse population, faces unique challenges in managing healthcare-associated outbreaks. This study aims to estimate the factors influencing the duration of infectious disease outbreaks in Toronto healthcare facilities, by examining various factors such as the type of healthcare setting, the causative agent, and the month when the outbreak occurred,

To better understand the relationships between these factors and outbreak duration, this study employs a Bayesian negative binomial regression model. This statistical model is well-suited to handling count data and predicting the number of days an outbreak lasts. By identifying the drivers of outbreak duration, the study will help inform strategies for improving outbreak preparedness and response. These findings are important for optimizing resource allocation, ensuring that healthcare facilities are well equipped to handle outbreaks, and improving response efforts.

@@

To examine the effects of factors in healthcare facilities in Toronto on outbreak duration, this report uses the Outbreaks in Toronto Healthcare Institutions dataset, which contains data from January 2016 to November 2024. The dataset is provided by Toronto Public Health, through City of Toronto Open Data Portal [@datasource]. Since no other datasets specific to healthcare institutions in Toronto are available, no additional data sources were considered. The dataset tracks reported outbreaks of gastroenteric and respiratory illnesses in Toronto healthcare institutions and contains detailed information on outbreak settings, causative agents, and outbreak durations. Following the principles from Telling Stories with Data [@textbook], we examine how the characteristics of outbreaks, such as the type of healthcare institution, the causative agent, and the month the outbreak began, influence their durations. A sample of the cleaned dataset is shown in @tbl-cleaned.

@@

### Outbreak setting
The Outbreak Setting variable is categorical and identifies the type of healthcare institution where the outbreak occurred, such as hospitals, long-term care homes (LTCH), or retirement homes. It provides insights into the environments most affected by outbreaks.

@@

### Causative agent
The Causative Agent variable is categorical and reflects the infectious agents responsible for outbreaks. While the original dataset contains 55 agents, they are grouped into seven broader categories to simplify the analysis and enhance interpretability.

@@

### Month outbreak began
The Month variable is categorical and records the calendar month when each outbreak started as a name (e.g., Jan, Feb). It reflects seasonal trends and potential patterns in infection rates. This variable is extracted from the date where each outbreak began from the original dataset and converted from a number to the corresponding month name.

@@

# Model {#sec-model}

## Model overview

@@

## Model setup

@@

## Model selection

@@

## Model diagnostics and validation

@@

# Results {#sec-results}

Our results are summarized in @tbl-model-summary using the modelsummary package [@modelsummary], where the model is tidied with the broom.mixed package [@broom] and the output table is scaled down with the kableExtra package [@kableExtra]. In the summary the model automatically selects the first category of each factor based on alphabetical orders of the category names. The intercept represents that the baseline outbreak duration is 2.257 days when all predictors are at their reference levels. For the outbreak setting, the reference category is "Hospital-Acute Care", all other outbreak settings are compared to "Hospital-Acute Care" to evaluate their influence on outbreak duration. The reference category is "Coronavirus" for the causative agent, the reference category is "April" for the month. By comparing the results, we can observe how each variable influences the duration differently.

@@

Based on the 90% credibility interval plot for the model in @fig-90-interval, we can analyze the influence of each predictor on outbreak duration:
Intercept: The intercept coefficient estimate lies above 0, indicating that the baseline outbreak duration (when all predictors are at their reference levels) is positively associated with longer outbreak durations. The confidence intervals for most variables are relatively small, with the largest interval reaching around 0.25. A wider interval suggests greater uncertainty and higher variation in the estimate for that predictor. Conversely, narrower intervals indicate more precise estimates of the effect of the predictor on outbreak duration.

Based on @fig-90-interval and @tbl-model-summary, we can see that for outbreak settings, the model suggests that outbreaks in Retirement Homes and Long-Term Care Homes (LTCHs) tend to last longer than other settings. This supports the expectation that these facilities, housing more vulnerable demographics, are more prone to prolonged outbreaks. On the other hand, Hospital-Psychiatric outbreaks tend to be shorter, as indicated by the negative coefficient. This supports the variability in duration across different healthcare settings. Regarding the causative agents, Respiratory Syncytial Virus (RSV), Parainfluenza, Norovirus, Enterovirus/Rhinovirus, and COVID-19, all show positive coefficient estimates, with COVID-19 being much larger compared to all other agents. This suggests that outbreaks caused by these agents last longer than those caused by Coronavirus (the reference group). In contrast, Influenza outbreaks are expected to last shorter, as indicated by its negative coefficient, while Metapneumovirus shows a marginal increase in duration, though its coefficient is very close to 0. Finally, the month of the outbreak also plays a role in its duration. The model suggests that outbreaks occurring in August and January are generally shorter compared to April (the reference month), while December outbreaks tend to last longer, which may suggest seasonal variation in outbreak duration. Overall, the 90% credibility intervals for the coefficients highlight the significance of the predictors in influencing outbreak duration, with some factors such as the type of healthcare setting and causative agent having a more substantial effect, while others, like month of occurrence, show moderate influences.

# Discussion {#sec-discussion}

## Long-term care homes and retirement homes are most susceptible to outbreaks
The type of healthcare setting where an outbreak occurs significantly impacts its duration. According to the model’s results, outbreaks in Long-Term Care Homes (LTCHs) and Retirement Homes are more likely to last longer compared to other settings. This finding aligns with the vulnerability of populations in these settings, particularly the elderly, who may have weakened immune systems or multiple preexisting conditions. This observation is consistent with a population-based study of Retirement Homes (RHs) in Ontario, which reported that RHs are highly susceptible to SARS-CoV-2 outbreaks due to home-level and community-level factors. Larger resident capacities, co-location with Long-Term Care facilities, and offering multiple services onsite were identified as significant contributors to outbreak risk. These factors suggest increased staff movement and staff-resident interactions may exacerbate outbreak dynamics in these settings [@retirementCovidRisks]. Furthermore, the study observed that RHs in regions with higher community-level ethnic concentration or increased regional COVID-19 incidence were at elevated risk of outbreaks, although the sensitivity to these factors diminished in Wave 2 due to improved preventative measures, such as mandatory surveillance testing and stricter infection control practices [@retirementCovidRisks].

The facilities of Long-Term Care Homes and Retirement Homes often lack the resources and infrastructure of hospital settings, which may contribute to longer outbreak durations in these environments. Previous research highlights staffing shortages in Long-Term Care Homes further worsen these challenges. The Long-Term Care workforce, has faced retention and recruitment issues for years. The COVID-19 pandemic intensified these problems, leading to burnout and workforce attrition despite wage increases and hiring incentives [@longTermCareStaffing]. Additionally, limited training and career advancement opportunities for direct care aides further strain care quality. These staffing issues, combined with high operational costs, hinder the ability of LTCH facilities to manage outbreaks effectively [@longTermCareStaffing]. Addressing these systemic challenges is critical to improving outbreak management in LTCH and Retirement Homes.

## The Impact of COVID-19 vs regular Coronavirus

The analysis shows that COVID-19 had a significant impact on the duration of outbreaks in healthcare settings, especially when compared to other respiratory pathogens like the common coronavirus. COVID-19 outbreaks were found to last much longer than those caused by regular coronaviruses, a finding consistent with the global experience during the pandemic. Several factors contributed to this prolonged duration, including the high transmissibility of the virus, its asymptomatic spread, and the challenges of managing outbreaks in healthcare settings housing vulnerable populations [@retirementCovidRisks]. Additionally, the virus’s evolving nature, with new variants emerging over time, further complicated efforts to contain it, requiring healthcare facilities to frequently adjust infection control measures.

While the model suggests that regular coronaviruses did not have a similar effect on outbreak duration, as seen in @tbl-model-summary, the coefficient is close to that of other causative agents. This indicates that once the COVID-19 pandemic is under control, coronaviruses may not pose the same level of challenge to healthcare systems. The findings highlight the importance of ongoing preparedness for future global health crises, where the emergence of new infectious diseases or highly transmissible variants could once again stretch healthcare resources. Continued investment in healthcare infrastructure, surveillance, and rapid response capabilities will be essential to minimize the potential impact of future pandemics on healthcare facilities, ensuring they are well-equipped to manage outbreaks of varying durations.

## The Effect of Winter on Outbreak Duration
The model suggests that outbreaks occurring in winter, particularly in December, tend to last longer compared to other months. This seasonal variation is likely due to higher rates of respiratory illnesses such as influenza and RSV during colder months, combined with holiday-related gatherings and staff shortages. The strain on healthcare facilities, particularly those housing vulnerable populations, can further extend outbreak durations. These findings highlight the need for targeted preparedness during winter, including increased staffing, early vaccinations, and robust infection control measures to minimize the impact of seasonal outbreaks.

## Weaknesses and Next Steps

One limitation of this analysis is the reliance on available outbreak data, which may not fully capture all outbreaks in healthcare settings. For instance, some incidents might have gone unreported, or data may have been misclassified, which could introduce bias or reduce the generalizability of the findings. Additionally, the model does not account for interactions between different factors, such as the combined effects of outbreak setting and causative agent, which could provide more nuanced insights. The assumptions made in the negative binomial regression model also rely on the availability of complete and accurate data, which may not always be the case, particularly with retrospective datasets. Furthermore, although the Mean Absolute Error (MAE) of 6.52 days is relatively small, it still represents a notable deviation from the actual outbreak duration, suggesting that further investigation and model refinement are needed to improve accuracy.

Future research could enhance this study by incorporating more detailed data, including healthcare interventions and patient information. Exploring the interaction between predictors like outbreak setting, causative agent, and seasonal trends could provide deeper insights into outbreak dynamics. The model also does not incorporate the potential influence of COVID-19 vaccinations on outbreak duration, an important factor that could have significantly altered the course of outbreaks, particularly in healthcare settings. Further research could explore the role of vaccination coverage and its interaction with other variables to improve predictive accuracy.

@@

## Model details

### Posterior predictive check

@@

### Leave-One-Out (LOO) Cross Validation (CV) Comparison

@@

### Model summary

@@

### Diagnostics
